---
title: "ENVS 193DS Final"
author: "Macey Hartmann"
date: "2026-03-17"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

[GitHub Repo](https://github.com/maceyhartmann/ENVS-193DS_winter-2026_final)

# Set up

```{r}
#| label: packages and data
#| warning: false

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection
library(DHARMa) # diagnostic check
library(rstatix) # cliffs delta
library(effsize) # add this — contains cliff.delta()
library(showtext) # custom fonts
library(systemfonts) # custom fonts

# reading in data

kelp <- read_csv(here("data", "Annual_Kelp_All_Years_20250903.csv")) # read in KFCD abundance and size of giant kelp data and save as an object called kelp
mopl <- read_csv(here("data", "MOPL_nest-site_selection_Duchardt_et_al._2019.csv")) # read in Mountain plover data and save as an object called mopl
my_data <- read_csv(here("data", "Personal_Data_ENVS194DS.csv")) # read in the personal data as an object named "my_data"
```

# Problem 1. Research writing

## a. Transparent statistical methods 

In part 1, they used a simple linear regression, because they are testing whether a single continuous predictor variable (precipitation) predicts a continuous response variable (flooded wetland area). In part 2, they used a Kruskal-Wallis test, because they are comparing the median of a continuous response variable (flooded wetland area) across more than two groups of a single categorical predictor variable (water year classification: wet, above normal, below normal, dry, critical drought).

## b. Figure needed

The coworker should make a jitter plot with an overlaid side by side box plot. The x-axis should display water year classification as a categorical variable with five groups (wet, above normal, below normal, dry, critical drought), and the y-axis should display flooded wetland area in m². The boxplot component should show the median as a horizontal line within the box, with the box itself representing the interquartile range and whiskers extending to the full range of the data as these elements directly correspond to the Kruskal-Wallis test's comparison of medians across groups. The jittered points overlaid on each box should show the underlying raw data for each water year classification, giving the reader a sense of spread and sample size within each group. Using different colors for each water year category would further help a non-statistical audience visually distinguish between the five groups.

## c. Suggestions for rewriting

In part 1, we did not find a significant relationship between precipitation and flooded wetland area in the San Joaquin River Delta (linear regression, F(df model, df residual) = test statistic, R² = coefficient of determination, p = 0.11, α = significance level). In part 2, we did not find a significant difference in median flooded wetland area across water year classifications (wet, above normal, below normal, dry, and critical drought) (Kruskal-Wallis test, $\chi$^2^(df) = test statistic, p = 0.12, α = significance level).

## d. Interpretation

One additional variable that could influence flooded wetland area is the number of water management interventions per land parcel (active reflooding events), which is a discrete numeric variable. This variable may influence wetland flooding because infrastructure upgrades and incentive programs that encouraged landowners to retain flooded wetlands longer into the breeding season were associated with the observed increases in flooded wetland area over time.

# Problem 2. Data visualization

## a. Cleaning and summarizing

```{r}
#| label: data-visualization
#| message: false
#| warning: false
kelp_clean <- kelp |> # create a new object called kelp_clean using kelp
  clean_names() |> # clean up column names so that only lower case letters and underscores are used
  filter(site %in% c("IVEE", "NAPL", "CARP")) |> # filter by site 
  filter(!(date == "2000-09-22" & transect == "6" & quad == "40")) |> # remove observation from 2000-09-22 at transect 6, quad 40
  # creating a new column called site_name and relabeling sites with full names
  mutate(site_name = case_when(
  site == "NAPL" ~ "Naples",
  site == "IVEE" ~ "Isla Vista",
  site == "CARP" ~ "Carpinteria" )) |>
  # converts site_name to a factor and sets level order to Naples, Isla Vista, Carpinteria for plotting
  mutate(site_name = as_factor(site_name),
       site_name = fct_relevel(site_name,
                               "Naples", "Isla Vista", "Carpinteria")) |>
  group_by(site_name, year) |> # group by site and year for summarizing
  summarize(mean_fronds = mean(fronds)) |> #  calculate mean fronds per site / year group
  ungroup() # remove grouping

slice_sample(kelp_clean, n = 5) # display 5 random rows
str(kelp_clean) # display structure
```

## b. Understanding the data

The observation from 2000-09-22 at transect 6, quad 40 has a frond count value of -99999, which is a value used to indicate missing data (either because data were not collected for that species or because a datasheet was lost, as noted in the Methods and protocols section of the data page). This value is filtered out because -99999 does not represent a real frond count and would skew any calculations of mean fronds if retained.

## c. Visualize the data

```{r}
#| echo: false
font_add_google("Lato", "lato")  # "Lato" is the font name, "lato" is the alias
showtext_auto()
```


```{r}
#| label: visualize-the-data
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 8
# create a plot with the cleaned kelp data set
ggplot(data = kelp_clean,
       aes(x = year, # x-axis dates
           y = mean_fronds, # y-axis mean fronds
           color = site_name)) + # different colors for each site 
  # first layer: lines
  geom_line() +
  # second layer: points
  geom_point() + 
  # coloring each site a new color
  scale_color_manual(values = c(
    "Naples" = "#4a9e9e",
    "Isla Vista" = "#c07a6d",
    "Carpinteria" = "#8a8a8a")) +
  labs(title = "Sites vary in mean kelp fronds per year", # graph title
       x = "Year", # x-axis label
       y = "Mean kelp fronds per year", # y-axis label
       color = "Site") + # capitalizing Site in key 
theme_light(base_family = "serif") +
theme(
  plot.title = element_text(hjust = 0.5, family = "serif"),
  axis.title = element_text(family = "serif"),
  axis.text = element_text(family = "serif"),
  legend.title = element_text(family = "serif"),
  legend.text = element_text(family = "serif"),
  panel.border = element_blank()
)
```

## d. Interpretation

The Isla Vista site shows the most variable mean kelp fronds per year, as evidenced by the dramatic peaks and troughs in its line. The Isla Vista site spikes to nearly 40 fronds around 2008 and again to ~38 fronds around 2022, then drops to near or below 5 fronds in several intervening years, producing the largest overall range and most erratic trajectory of the three sites. 

The Carpinteria site shows the least variable mean kelp fronds per year, as its line remains relatively flat and close to the bottom of the plot for the entire time series, generally fluctuating between approximately 3 and 13 fronds per year with no dramatic spikes, indicating a much smaller and more consistent range compared to the other two sites.

# Problem 3. Data analysis

## a. Response variable 

In this data set, a 1 indicates that a nest site was actively used by a Mountain Plover for nesting. A 0 indicates an unused site, specifically a randomly selected point within the same colony where no plovers were detected and no nests were observed during surveys in that year.

## b. Purpose of study

Visual obstruction (vo_nest) is a numeric continuous variable measuring the height at which vegetation is sufficiently dense to obscure a 2.5-cm wide pole when viewed from a 1-m height, incorporating both vegetation height and density. As visual obstruction increases, the probability of Mountain Plover nest site use would decrease because Mountain Plovers prefer sparse, short vegetation that allows them to detect approaching ground predators while still providing some concealment for a sitting bird. This information was found in paragraph 4 of the discussion section. 

Distance to prairie dog colony edge (edgedist) is a continuous variable measuring how far a given point is from the nearest edge of the prairie dog colony. As distance to colony edge increases, the probability of Mountain Plover nest site use would decrease because plovers tend to nest closer to colony edges where complementary resources such as taller vegetation for thermoregulation and predator cover are more accessible for adults traveling with broods. This information was found in paragraph 2 of the discussion section. 

## c. Table of models 

| Model number | Visual obstruction | Distance to prairie dog colony | Model description             | 
|--------------|:------------------:|:------------------------------:|:-----------------------------:| 
| 0            |                    |                                | No predictors (null model)    |
| 1            |         X          |               X                | All predictors (full model)   | 
| 2            |                    |               X                | Distance to colony edge only  |   
| 3            |         X          |                                | Visual obstruction only.      |    

## d. Run the models

```{r}
#| warning: false
mopl_clean <- mopl |> # create a new object called mopl_clean using mopl
  clean_names() |> # clean up column names so that only lower case letters and underscores are used
  # create binary response variable: 1 = used nest site, 0 = available site
  mutate(used_bin = case_when(
    id == "used" ~ 1,
    id == "available" ~ 0
  )) |>
  # keep only the response variable and two predictors of interest
  select(used_bin, edgedist, vo_nest)

# model 0: null model, no predictors
mopl_model0 <- glm(used_bin ~ 1, # response
                   data = mopl_clean, # data frame
                   family = "binomial") # error distribution

# model 1: saturated model, both visual obstruction and distance to colony edge
mopl_model1 <- glm(used_bin ~ vo_nest + edgedist,  # response vs. edge distance and visual obstruction
              data = mopl_clean, # data frame
              family = "binomial") # error distribution

# model 2: distance to colony edge only
mopl_model2 <- glm( used_bin ~ edgedist, # response vs. edge distance
                    data = mopl_clean, # data frame
                     family = "binomial") # error distribution

# model 3: visual obstruction only
mopl_model3 <- glm( used_bin ~ vo_nest, # response vs. visual obstruction
                    data = mopl_clean, # data frame
                     family = "binomial") # error distribution
```

## e. Select the best model

```{r}
#| warning: false
# calculate the AIC for each model
AIC( 
  mopl_model0,
  mopl_model1,
  mopl_model2,
  mopl_model3
) 
```

The best model as determined by Akaike's Information Criterion (AIC) is the model predicting the probability of Mountain Plover nest site occupancy using visual obstruction as the sole predictor (AIC = 331.77). This model had a lower AIC value than the saturated model containing both visual obstruction and distance to prairie dog colony edge, the model containing only distance to colony edge, and the null model.

## f. Check diagnostics

```{r}
#| warning: false
plot(
  simulateResiduals(mopl_model3) # check DHARMa diagnostic plots for mopl_model3
)
```

Simulated residuals from the DHARMa package indicate that model assumptions are met. The QQ plot shows points closely following the expected diagonal with no significant deviation (KS test: p = 0.72, dispersion test: p = 0.99, outlier test: p = 1), and the residuals vs. predicted plot shows no notable patterns, suggesting the model is appropriately specified. 

## g. Visualize model predictions

```{r}
#| warning: false
# predicted probabilities from model 3
preds <- ggpredict(
  mopl_model3,
  terms = ("vo_nest [all]")
) 
# base layer: ggplot
ggplot(mopl_clean, 
       aes(x = vo_nest, # x-axis visual obstruction
           y = used_bin) # y-axis binary nest site use
       ) + 
  # first layer: points
  geom_point(size = 4,
             color = "forestgreen", 
             alpha = 0.2) + 
  # second layer: add 95% CI around the predictions and the underlying data
  geom_ribbon(data = preds,
              aes(x = x, 
                  y = predicted, 
                  ymin = conf.low, 
                  ymax = conf.high),
              alpha = 0.3, # transparency
              fill = "#c07a6d") + # custom color for ribbon)
  # third layer: model predictions line
  geom_line(data = preds, 
            aes(x = x, 
                y = predicted), 
                color = "#c07a6d") + # custom color for line
  scale_y_continuous(limits = c(0,1),
                     breaks = c(0,1)) +
  labs(x = "Visual Obstruction (cm)", 
       y = "Probability of Nest Site Occupancy (1 = Yes, 0 = No)") +
 # remove gridlines and apply clean theme
  theme_light() +
  theme(panel.grid = element_blank()) # remove any remaining gridlines
```

## h. Write a caption for your figure 

**Figure 1. Predicted probability of Mountain Plover nest site occupancy decreases with increasing visual obstruction.** The predicted probability of nest site occupancy (y-axis) is shown as a function of visual obstruction measured in centimeters (cm) on the x-axis, based on a logistic regression model. The red line represents the model-predicted probability of nest site occupancy, and the shaded ribbon represents the 95% confidence interval around the predictions. Points along y = 1 and y = 0 represent the underlying observed data for used nest sites and unused available sites, respectively, with transparency (alpha) indicating overlapping observations. Data citation: Duchardt, Courtney; Beck, Jeffrey; Augustine, David (2020). Data from: Mountain Plover habitat selection and nest survival in relation to weather variability and spatial attributes of Black-tailed Prairie Dog disturbance [Dataset]. Dryad. https://doi.org/10.5061/dryad.ttdz08kt7.

## i. Calculate model predictions

```{r}
#| warning: false
ggpredict(mopl_model3, # use model 3 (visual obstruction only)
          terms = "vo_nest [0]") # visual obstruction = 0
```

## j. Interpret your results

The best model as determined by AIC included only visual obstruction as a predictor of Mountain Plover nest site occupancy, suggesting that distance to prairie dog colony edge did not meaningfully improve model fit beyond visual obstruction alone. At zero visual obstruction (completely bare, unobstructed ground), the predicted probability of nest site occupancy is 0.73 or 73% (95% CI: 0.65, 0.80), and as shown in Figure 1, this probability declined steeply and nonlinearly as visual obstruction increased, approaching near-zero probability at visual obstruction values above approximately 6 cm. This negative relationship between visual obstruction and nest site occupancy is biologically consistent with Mountain Plover habitat preferences described in the discussion, where the authors note that plovers select nest sites with sparse, short vegetation in the range of 3–5 cm because the short, dense layer provides some concealment for a sitting bird while still allowing the plover to scan for and detect approaching ground predators. Additionally, Mountain Plovers blend in especially well with bare soil when viewed by aerial predators, further explaining the strong preference for low visual obstruction at nest sites. These findings reinforce the critical role of black-tailed prairie dogs as ecosystem engineers that maintain the open, disturbed habitat structure that Mountain Plovers require for successful nesting.

# Problem 4. Affective and exploratory visualizations

## a. Comparing visualizations

**How are the visualizations different?**
The affective visualization represents reading data through a hand-drawn tree where each branch represents a reading location (campus, home, coffee shop, outside, other), each leaf represents an individual reading session, and leaf color encodes pages read, making the data experiential and emotionally resonant rather than statistically precise. In contrast, the exploratory visualizations use conventional statistical graphics (a scatterplot showing the relationship between reading duration and pages read, and a boxplot comparing pages read across reading locations). The exploratory visualization prioritizes accurate representation of distributions and relationships while the affective visualization shows aesthetic, emotions, and creativity. 

**What similarities do you see?**
Both visualizations have the same core variables from the personal dataset: reading location and pages read. The use of color to distinguish categories or encode magnitude appears across both the affective visualization (leaf colors representing pages read) and the exploratory boxplot (different colors per location).

**What patterns do you see?**
The exploratory boxplot suggests campus reading sessions tend to produce more pages read than home sessions, and the scatterplot shows a positive relationship between reading duration and pages read. These same patterns are present but harder to read precisely in the affective visualization, where the home branch appears to have more leaves clustered at warmer colors (higher pages) than the campus branch, though the non-standard format makes direct comparison less intuitive.

**Feedback and implementation**
During office hours, feedback from my peers suggested making the leaf size or color scale more explicit so viewers could decode the data without a legend being necessary. I tried to implement this by adding a more consistent color gradient across leaves in the final affective visualization and using a darker yellow color. I also incorporated newspaper into the whole visualization instead of just the tree base to play around with more of a mixed media and get more creative. 

## b. Designing an analysis

To examine whether the day type (school day or non-school day) predicts the number of pages read per session, I would propose a Wilcoxon rank sum test. My response variable, pages read, is a continuous numeric variable, and my predictor variable, day type, is a categorical variable with two groups (school day and non-school day). A Wilcoxon rank sum test is appropriate here because it compares medians between two groups of a categorical predictor, which directly addresses my research question of whether the type of day influences how much I read. Additionally, given the small sample size in my personal dataset and visual deviations observed in the QQ plots, the data are unlikely to meet the normality assumption required for a two-sample t-test, making the Wilcoxon rank sum test a more appropriate non-parametric alternative that does not assume a normal distribution.

## c. Check any assumptions and run your analysis

### Check all assumptions 

The key assumptions for a Wilcoxon rank sum test include that the response variable is continuous (met by nature of the variable, as pages read is a continuous numeric variable), the predictor variable has two categorical, independent groups (met by nature of the variable (school day or non-school day),  observations are independent (met by study design, as each reading session is a separate, independent observation), and that the two groups have similar distribution shapes (must do visual assumption check). 

```{r}
#| echo: false
#| warning: false
my_clean_data <- my_data |> # create a new object called my_clean_data using my_data
  clean_names() |> # clean up column names so that only lower case letters and underscores are used
  mutate(school_day = recode(school_day,
                             "yes" = "School Day",
                             "no" = "Non-School Day"))
```

```{r}
#| warning: false
# base layer: ggplot
ggplot(data = my_clean_data, # use cleaned personal data 
       aes(x = pages_read, # pages read on x-axis
           fill = school_day)) + # fill bars by school day 
 # first layer: histogram
   geom_histogram(binwidth = 10, 
                 alpha = 0.7, # transparent bars 
                 color = "white") + # white outline
  facet_wrap(~school_day) + # separate panel by day type
  scale_x_continuous(breaks = seq(0, 80, by = 10)) + # whole number x-axis ticks
  scale_fill_manual(values = c("Non-School Day" = "#378ADD", # blue for non-school day
                               "School Day" = "#D85A30")) + # red for school day
  labs(title = "Distribution of Pages Read by Day Type", # title 
       x = "Pages Read", # x-axis title
       y = "Number of Reading Sessions") + # y-axis title
  theme_linedraw() + # change theme from default
  theme(legend.position = "none", # remove legend 
  panel.grid.minor = element_blank(), # remove minor gridlines 
        axis.line.x = element_line(color = "black")) 
```

Both distributions are unimodal and roughly symmetric, with non-school days centered around 35–40 pages and school days centered around 20–25 pages, suggesting that the similar shape assumption of the Wilcoxon rank sum test is reasonably satisfied.

### Wilcoxon rank sum test

#### Summary of median and n values
```{r}
#| warning: false
my_clean_data |>
  group_by(school_day) |> # group by school day vs non school day
  summarize(
    median = median(pages_read), # calculate median pages read
    n = n() # count number of reading sessions per group
  )
```

#### Run a Wilcoxon rank sum test 

```{r}
#| warning: false
wilcox.test(pages_read ~ school_day, # compare pages read by day type
            data = my_clean_data) # use cleaned personal data
```

#### Calculate effect size

```{r}
# calculate Cliff's delta (effect size)
cliff.delta( 
  pages_read ~ school_day, # response ~ predictor
  data = my_clean_data # use personal data
)
```

I found a small (Cliff's δ = 0.32) difference in median pages read between non-school days (n = 12) and school days (n = 22). (Wilcoxon rank-sum test, W = 174.5, p = 0.13, ⍺ = 0.05). A continuity correction was applied due to ties in the data. Pages read per session tended to be higher on non-school days (median = 37.5) than on school days (median = 28).

## d. Create a visualization

```{r}
# base layer: ggplot
ggplot(data = my_clean_data,
       aes(x = school_day, # categorical predictor on x-axis
           y = pages_read)) + # response variable on y-axis
  
  # first layer: boxplot showing median and spread
  geom_boxplot(aes(fill = school_day), # fill boxplot by day type
               width = 0.6,
               outliers = FALSE) + # avoid outliers being shown twice
  
  # second layer: jittered raw data points
  geom_jitter(width = 0.1, # horizontal spread
              alpha = 0.4,  # set transparency of points
              height = 0) + # no vertical jitter
  
  scale_y_continuous(breaks = seq(0, 80, by = 10)) + # y-axis breaks every 10 pages
  
  # custom colors 
 scale_fill_manual(values = c(
    "Non-School Day" = "lightblue", # teal for non-school day
    "School Day" = "#FFCCCB")) + # purple for school day
  
  # descriptive axis labels with units
  labs(x = "Day Type", # x-axis title
       y = "Pages Read (pages per session)") + # y-axis title
  
  theme_classic() + # changed from default
  theme(legend.position = "none") # remove redundant legend

```

## e. Write a caption

**Figure 1. Pages read per session by day type.** Boxplots show the median (center line), interquartile range (box), and 1.5× IQR (whiskers) for non-school days (blue, n = 12) and school days (pink, n = 22). Points represent individual reading sessions. Non-school days show a higher median pages read (37.5 pages) compared to school days (28 pages).

## f. Write about your results

I did not find a significant difference in median pages read between non-school days (n = 12, median = 37.5) and school days (n = 22, median = 28) (Wilcoxon rank-sum test: W = 174.5, p = 0.13, α = 0.05), with a small effect size (Cliff's δ = 0.32). Although non-school days had a median 9.5 pages higher than school days, the p-value exceeds the significance threshold and the confidence interval for Cliff's δ spans zero (-0.10, 0.65), suggesting this difference could be due to random chance. Day type does not appear to have a significant influence on the number of pages I read per session.

## g. Sharing your affective visualization

I participating in Week 10 section by sharing my affective visualization. 


